---
title: "analyze_butterfly_activity"
author: "Jacob Idec"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1: Which activity metrics are most distinctive? Can we find activity metrics that have a phylogenetic signature?

```{r physig, results='hide'}
library(colorEvoHelpers)
library(Momocs)
sscs <- addHrFreqsToSSCs(sscs,start_trim=6,end_trim=6,normalize=T,scale=T)
hr_freqs_shp <- sscs$hr_freqs
for(i in 1:length(hr_freqs_shp)){
    curve_mat <- cbind(as.numeric(names(hr_freqs_shp[[i]])),hr_freqs_shp[[i]])
    curve_mat[,2] <- curve_mat[,2] /  max(curve_mat[,2])
    curve_mat[,2] <- scale(curve_mat[,2],center=TRUE)
    hr_freqs_shp[[i]] <- curve_mat
    rownames(hr_freqs_shp[[i]]) <- NULL
}
shps <- Momocs::Opn(hr_freqs_shp)
op <- opoly(shps, degree=5)
op.p <- PCA(op)
plot(op.p,cex=0.3)
op.p$fac
?plot.PCA

sscs$shp_pc1 <- op.p$x[,1]
sscs$shp_pc2 <- op.p$x[,2]

# summarize to species for phylogenetic analysis
library(dplyr)
sscs_sp <- sscs %>%
    group_by(species) %>%
    dplyr::summarise(q5=mean(q5,na.rm=T),q50=mean(q50,na.rm=T),q95=mean(q95,na.rm=T),
                     diff1=mean(diff_third1_mean,na.rm=T),
                     diff2=mean(diff_third2_mean,na.rm=T),
                     diff3=mean(diff_third3_mean,na.rm=T),
                     pc1=mean(pc1,na.rm=T),pc2=mean(pc2,na.rm=T),
                     freq_pc1=mean(freq_pc1,na.rm=T),freq_pc2=mean(freq_pc2,na.rm=T),
                     freq_pc3=mean(freq_pc3,na.rm=T),freq_pc4=mean(freq_pc4,na.rm=T),
                     freq_pc5=mean(freq_pc5,na.rm=T),
                     shp_pc1 = mean(shp_pc1,na.rm=T),shp_pc2 = mean(shp_pc2,na.rm=T),
                     wingspan=mean(wingspan,na.rm=T), mean_n_obs=mean(obs_n,na.rm=T), n_cells=n(),
                     tmax=mean(tmax,na.rm=T),daylength=mean(daylength,na.rm=T))
sscs_sp$clade <- sscs_sp$species
sscs_sp_10 <- dplyr::filter(sscs_sp,n_cells > 10)
sscs_sp_20 <- dplyr::filter(sscs_sp,n_cells > 20)

# pcs
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc1",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc2",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc3",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc4",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc5",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

# onsets/offsets
assessTraitsOnTree(df=sscs_sp_20,trait_colname="q5",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="q50",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="q95",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

# shape pcs
assessTraitsOnTree(df=sscs_sp_20,trait_colname="shp_pc1",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="shp_pc2",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

# are pcs2 cor with predictors?
describeCols(sscs,colnames=c("shp_pc2","freq_pc4", "wingspan","tmax","daylength","obs_n"),corrplot_only = T)

# are predictors conserved?
sscs_sp_20$n_cells <- as.numeric(sscs_sp_20$n_cells)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="n_cells",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="tmax",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="daylength",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

```

PC4 has signal

q5, q50, and q95 kind of are but not nearly as much as PC1

## Q2: What is the best model formula by AIC removing multicollinear vars?

```{r cars}

library(colorEvoHelpers)

vifsAndStep(sscs,response="shp_pc2",formula="tmax + daylength + vp + precip + srad + 
                                                  + lat + wingspan",mixed=FALSE)

vifsAndStep(sscs,response="shp_pc2",formula="tmax + daylength + precip + srad + 
                                                  + lat",mixed=FALSE)
vifsAndStep(sscs,response="shp_pc2",formula="tmax + daylength + precip +
                                                  + lat",mixed=FALSE)

# swap lat to tmax bc they are correlated and tmax is more interesting
getPlotLM(sscs,model_type="lmer",response="shp_pc2",formula="tmax + daylength + precip + wingspan + obs_n + (1|species)")

getPlotLM(sscs,model_type="lmer",response="freq_pc1",formula="tmax + daylength + precip +
                                                  + (1|species) + (1|grid_id)")
getPlotLM(sscs,model_type="lmer",response="freq_pc2",formula="tmax + daylength + precip +
                                                  + (1|species) + (1|grid_id)")
getPlotLM(sscs,model_type="lmer",response="shp_pc2",formula="tmax + daylength + precip + wingspan +
                                                  + (1|species)")

# try pglmm
getPlotLM(sscs_phytrim,model_type="pglmm",
          phy=sscs_phy,
          response="q50",formula="tmax + daylength + precip + lat + (1|species__)")
```

## Occupancy modeling?

How can we fit this into an occupancy modeling framework?

each site is one hour interval in an SSC

site level covariates - temperature, precipitation, daylength, wingspan

biological classes -

observation level covariates - intensity of the observer baseline in the hour interval SSC

would give probability of occupancy at each sp cell and time period

ask rob how long occupancy is gonna fit
