---
title: "analyze_butterfly_activity"
author: "Jacob Idec"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1: Which activity metrics are most distinctive? Do activities have a phylogenetic signature?

```{r physig, results='hide'}
library(colorEvoHelpers)

source("src/addHrFreqsToSSCs.R")
sscs <- addHrFreqsToSSCs(sscs,start_trim=6,end_trim=6,normalize=F,scale=T)

# summarize to species for phylogenetic analysis
library(dplyr)
sscs_sp <- sscs %>%
    group_by(species) %>%
    dplyr::summarise(q5=mean(q5,na.rm=T),q50=mean(q50,na.rm=T),q95=mean(q95,na.rm=T),
                     diff1=mean(diff_third1_mean,na.rm=T),
                     diff2=mean(diff_third2_mean,na.rm=T),
                     diff3=mean(diff_third3_mean,na.rm=T),
                     pc1=mean(pc1,na.rm=T),pc2=mean(pc2,na.rm=T),
                     freq_pc1=mean(freq_pc1,na.rm=T),freq_pc2=mean(freq_pc2,na.rm=T),
                     freq_pc3=mean(freq_pc3,na.rm=T),freq_pc4=mean(freq_pc4,na.rm=T),
                     freq_pc5=mean(freq_pc5,na.rm=T),
                     shp_pc1 = mean(shp_pc1,na.rm=T),shp_pc2 = mean(shp_pc2,na.rm=T),
                     wingspan=mean(wingspan,na.rm=T), mean_n_obs=mean(obs_n,na.rm=T), n_cells=n())

sscs_sp$clade <- sscs_sp$species

# get sp thresholds
sscs_sp_10 <- dplyr::filter(sscs_sp,n_cells > 10)
sscs_sp_20 <- dplyr::filter(sscs_sp,n_cells > 20)

cor.test(sscs$freq_pc1,sscs$obs_n)
# pcs
assessTraitsOnTree(df=sscs_sp_10,trait_colname="freq_pc1",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc2",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc3",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc4",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc5",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

# onsets/offsets
assessTraitsOnTree(df=sscs_sp_20,trait_colname="q5",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="q50",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="q95",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

# shape pcs
assessTraitsOnTree(df=sscs_sp_20,trait_colname="shp_pc1",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_10,trait_colname="shp_pc2",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

# is shp pc2 cor with predictors?
cor.test(sscs$shp_pc2,sscs$obs_n)
cor.test(sscs$shp_pc2,sscs$tmax)
cor.test(sscs$shp_pc2,sscs$daylength)

# are predictors conserved?
assessTraitsOnTree(df=sscs_sp_20,trait_colname="n_cells",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

assessTraitsOnTree(df=sscs_sp_20,trait_colname="tmax",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="q50",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="q95",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

# Hesperidae clearly has the most variation (evolution?) in activity
# remove it and see if phylosigs become sig
# need to load in iNat data for now bc family is not saved at present
library(readr)
bfly_obs <- read_delim("data/inat_obs_raw/usa_butterflies_inat_gbif.csv",delim = "\t", escape_double = FALSE, trim_ws = TRUE)
hesp_sp <- unique(filter(bfly_obs,family=="Hesperiidae")$species)
sscs_sp_10_rm_hesp <- dplyr::filter(sscs_sp_10,!(species %in% hesp_sp))

assessTraitsOnTree(df=sscs_sp_10_rm_hesp,trait_colname="pc1",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_10_rm_hesp,trait_colname="pc2",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

# is wingspan conserved?
assessTraitsOnTree(df=sscs_sp,trait_colname="wingspan",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

describeCols(sscs,colnames=c("shp_pc2","wingspan","tmax","daylength","obs_n"),corrplot_only = T)

sscs_sp_20$freq_pc4_inv <- -1 * sscs_sp_20$freq_pc4
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc4_inv",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

assessTraitsOnTree(df=sscs_sp_20,trait_colname="shp_pc2",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

```

PC4 has signal

q5, q50, and q95 kind of are but not nearly as much as PC1

## Q1: What is the best model formula by AIC removing multicollinear vars?

```{r cars}

library(colorEvoHelpers)

vifsAndStep(sscs,response="q5",formula="tmax + daylength + vp + precip + srad + 
                                                  + lat + wingspan",mixed=FALSE)

vifsAndStep(sscs_8s_rmbf_250km_30nobs,response="q5",formula="tmax + daylength + precip + srad + 
                                                  + lat",mixed=FALSE)
vifsAndStep(sscs_8s_rmbf_250km_30nobs,response="q5",formula="tmax + daylength + precip +
                                                  + lat",mixed=FALSE)

# swap lat to tmax bc they are correlated and tmax is more interesting
getPlotLM(sscs,model_type="lmer",response="q5",formula="tmax + daylength + precip +
                                                  + (1|species) + (1|grid_id)")
getPlotLM(sscs,model_type="lmer",response="q5",formula="tmax + daylength + precip +
                                                  + (1|species)")
getPlotLM(sscs,model_type="lmer",response="q50",formula="tmax + daylength + precip +
                                                  + (1|species)")
getPlotLM(sscs,model_type="lmer",response="q95",formula="tmax + daylength + precip +
                                                  + (1|species)")

getPlotLM(sscs,model_type="lmer",response="freq_pc1",formula="tmax + daylength + precip +
                                                  + (1|species) + (1|grid_id)")
getPlotLM(sscs,model_type="lmer",response="freq_pc2",formula="tmax + daylength + precip +
                                                  + (1|species) + (1|grid_id)")
getPlotLM(sscs,model_type="lmer",response="shp_pc2",formula="tmax + daylength + precip + wingspan +
                                                  + (1|species)")

# try pglmm
getPlotLM(sscs_phytrim,model_type="pglmm",
          phy=sscs_phy,
          response="q50",formula="tmax + daylength + precip + lat + (1|species__)")
```

## Q4: Do activities have a phylogenetic signature?

```{r physig, results='hide'}
library(colorEvoHelpers)

# pcs
sscs_sp_10$freq_pc1_inv <- -1 * sscs_sp_10$freq_pc1
assessTraitsOnTree(df=sscs_sp_10,trait_colname="freq_pc1_inv",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_10,trait_colname="freq_pc2",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

# try 20 cutoff
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc1",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc2",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

# onsets/offsets
assessTraitsOnTree(df=sscs_sp_10,trait_colname="q5",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_10,trait_colname="q50",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_10,trait_colname="q95",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

# Hesperidae clearly has the most variation (evolution?) in activity
# remove it and see if phylosigs become sig
# need to load in iNat data for now bc family is not saved at present
library(readr)
bfly_obs <- read_delim("data/inat_obs_raw/usa_butterflies_inat_gbif.csv",delim = "\t", escape_double = FALSE, trim_ws = TRUE)
hesp_sp <- unique(filter(bfly_obs,family=="Hesperiidae")$species)
sscs_sp_10_rm_hesp <- dplyr::filter(sscs_sp_10,!(species %in% hesp_sp))

assessTraitsOnTree(df=sscs_sp_10_rm_hesp,trait_colname="pc1",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_10_rm_hesp,trait_colname="pc2",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

# is wingspan conserved?
assessTraitsOnTree(df=sscs_sp,trait_colname="wingspan",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

```

PC1 is EXTREMELY phy cor while PC2 is not AT ALL

q5, q50, and q95 kind of are but not nearly as much as PC1

## Q3: What is the best model formula by AIC removing multicollinear vars for the curve PCs?

```{r cars}

library(colorEvoHelpers)

vifsAndStep(sscs,response="freq_pc1",formula="tmax + daylength + vp + precip + srad + 
                                                  + lat + wingspan",mixed=FALSE)
vifsAndStep(sscs,response="freq_pc2",formula="tmax + daylength + vp + precip + srad + 
                                                  + lat + wingspan",mixed=FALSE)

getPlotLM(sscs,model_type="lmer",response="freq_pc1",formula="tmax + daylength + wingspan
                                                  + (1|species)")
getPlotLM(sscs,model_type="lmer",response="freq_pc2",formula="tmax + daylength + precip
                                                  + (1|species)")

# with grid id
getPlotLM(sscs,model_type="lmer",response="freq_pc1",formula="tmax + daylength + precip + wingspan
                                                  + (1|species) + (1|grid_id)")
getPlotLM(sscs,model_type="lmer",response="freq_pc2",formula="tmax + daylength + wingspan
                                                  + (1|species)")

# try pglmm
m <- getPlotLM(sscs,
          tree_location = "data/misc/bf_species_tree.txt",remove_first_split=T,
          model_type = "pglmm",bayes=F,
          response="freq_pc1",formula="tmax + daylength + wingspan + (1 | species__)")

m <- getPlotLM(sscs,
          tree_location = "data/misc/bf_species_tree.txt",remove_first_split=T,
          model_type = "pglmm",bayes=F,
          response="freq_pc2",formula="tmax + daylength + precip + (1 | species__)")

```

## Q4: What are some possible issues with PC1? Is it related to sampling effort?

```{r pc1_effort}

library(colorEvoHelpers)

sscs <- addHrFreqsToSSCs(sscs,start_trim=5,end_trim=5,normalize = T,scale=T)
cor.test(sscs$wingspan,sscs$obs_n)
cor.test(sscs_sp_10$wingspan,sscs_sp_10$n)

cor.test(sscs$freq_pc1,sscs$obs_n)
cor.test(sscs_sp_10$freq_pc1,sscs_sp_10$n)

# investigate pc3 and 4 
assessTraitsOnTree(df=sscs_sp_10,trait_colname="freq_pc1",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_10,trait_colname="freq_pc2",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_10,trait_colname="freq_pc3",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_10,trait_colname="freq_pc4",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_10,trait_colname="freq_pc5",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

# investigate pc3 and 4 
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc1",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc2",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc3",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc4",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc5",tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

```

At the level, wingspan is only slightly correlated with sample size

At the level of species SSCs, wingspan is more heavily correlated 0.35 with the number of SSCs, but not crazy

However, there is a CRAZY cor of 0.99 between pc1 and the number of obs in an SSC
