---
title: "describe_butterfly_activity"
author: "Jacob Idec"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1: How is each response variable (metrics for when butterflies stop and start foraging or are most active) distributed and correlated?

```{r responses, echo=FALSE}

library(colorEvoHelpers)
sscs <- sscs_8s_rmbf_250km_30nobs
describeCols(sscs,colnames=c("q5","q50","q95",
                             "diff_third1_mean","diff_third2_mean","diff_third3_mean",
                             "freq_pc1","freq_pc2","freq_pc3","freq_pc4"))

```

q5, q50, and q95 are normally distributed

Some diffs are normal, others are tailed

q50 is VERY correlated with q5 which is interesting and makes sense- as onset increases the peak also increases

q50 is also cor with q95 - when peak is later, end is later

Clear weird difference between q95 and diff third 3 - negatively correlated

Clearly the weib and diff metrics mean very different things as they are not closely correlated

Freq_pc1 is heavily cor with q5, q50, and q95, which makes perfect sense bc it is a shift of activity later in the day

## Q2: How is each predictor variable distributed and correlated?

```{r preds, echo=FALSE}
describeCols(sscs,colnames=c("tmax","daylength","vp","precip","srad",
                                                  "lat","wingspan"))

describeCols(sscs,colnames=c("species"),factor=TRUE)

```

Daylength and precip are pretty weird looking

Vp extremely correlated with tmax and somewhat with precip

Srad also correlated with daylength

The vast majority of species represented have few SSCs, 40% of species have 3 or less SSCs

Looking at this, we probably want to drop vp and srad due to being correlated with others

## Q3: How are predictors and responses associated?

```{r p_r, echo=FALSE}

describeCols(sscs,colnames=c("q5","q50","q95",
                             "diff_third1_mean","diff_third2_mean","diff_third3_mean",
                             "freq_pc1","freq_pc2","freq_pc3",
                              "tmax","daylength","vp","precip","srad",
                                                  "lat","wingspan"),corrplot_only = T)
describeDfAssoc(sscs,colnames=c("q5","q50","q95",
                             "diff_third1_mean","diff_third2_mean","diff_third3_mean",
                              "tmax","daylength","vp","precip","srad",
                                                  "lat","wingspan"))

```

Spot trends we might look more carefully at in models

Freq_pc1 is heavily cor with srad - more srad means more obs earlier

## Q4: how are individuals spatially distributed?

```{r spatial_heatmap, echo=FALSE}
sscs$latitude <- sscs$lat
sscs$longitude <- sscs$lon
plotLatLonHeatmap(sscs) # need to add lons in data prep first!! 
```

Actually not that bad

## Q5: What do PCAs of the responses look like i.e how is the variation structured? What do the hour frequency PCs look like?

```{r response_pca, echo=FALSE}

pcdata <- runPlotBindPCA(na.omit(sscs),c("q5","q95","q50"))
sscs <- addHrFreqsToSSCs(sscs)
```

For the PCA of quantiles, the big cloud is 0 for PC2 and slightly less than 0 for PC1 ( later mid and later onset)

They they slowly disperse out to be postiive on PC1 (earlier mid and onset) , with a lot of variation in PC2 (later offset and earlier onset)

Plotting the PCs of the WHOLE curves is pretty amazing though, with 77% of the variation in the curves being in 1 PC representing less foraging activity in the midday, the inverse can be interpreted as more activity in the midday!!!

The 2nd PC explains the next 22%, and is basically a skew towards peaking earlier or later in the day!

Later in analysis we will find that these are more phylogenetically conserved than other metrics

## Q6: Is PC1 (in PCA without normalization) the baseline?

```{r bl_pca, echo=FALSE}

# plot insect obs
library(readr)
insect_obs <- read_delim("data/inat_obs_raw/usa_insects_inat_gbif.csv",delim = "\t", escape_double = FALSE, trim_ws = TRUE)
source("src/datetimeToSolarAndSeason.R")
insect_obs <- datetimeToSolarAndSeason(insect_obs,season_intervals_per_year = 8)
hist(insect_obs$local_hour[insect_obs$local_hour > 5 & insect_obs$local_hour < 19])

# plot bfly obs
bfly_obs <- read_delim("data/inat_obs_raw/usa_butterflies_inat_gbif.csv",delim = "\t", escape_double = FALSE, trim_ws = TRUE)
source("src/datetimeToSolarAndSeason.R")
bfly_obs <- datetimeToSolarAndSeason(bfly_obs,season_intervals_per_year = 8)
hist(bfly_obs$local_hour[bfly_obs$local_hour > 5 & bfly_obs$local_hour < 19])
hist(insect_obs$local_hour[insect_obs$local_hour > 5 & insect_obs$local_hour < 19])

source("src/addHrFreqsToSSCs.R")
sscs <- addHrFreqsToSSCs(sscs,start_trim=7,end_trim=6)

# one more phylo
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc3",trait_colname2="freq_pc1", tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

cor.test(sscs_sp$freq_pc4,sscs_sp$mean_n_obs)
#0.8048  0.1255 0.02303 0.01694 0.00854

#plot(sscs$freq_pc1,sscs$))

assessTraitsOnTree(df=sscs_sp_10,trait_colname="freq_pc3", tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)

cor.test(sscs$freq_pc3,sscs$freq_pc2)

sscs <- addHrFreqsToSSCs(sscs,start_trim=7,end_trim=6)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc1", tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc2", tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc3", tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc4", tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)
assessTraitsOnTree(df=sscs_sp_20,trait_colname="freq_pc5", tree_location="data/misc/bf_species_tree.txt",remove_first_split=T,legend=T)


# plot pc3 data for a particular species 
# see what score for a ssc is on pc1, then look for species with high or low value for pc1, do the same thing for pc2 

# everything like pc1 (0) are things that look like the observation effort curve 
# currently if you're at the 0 value 

cor.test(sscs$freq_pc1,sscs$daylength)

cor.test(sscs$tmax,sscs$wingspan)
```

PC1 looks very similar to both the insect baseline and the bfly baseline

It is the mean foraging curve that gets weighted up and down by the number of observations

## Q7: Does SVD and eigenshape analysis show the same results as PCA?

```{r preds, echo=FALSE}

# visualize pcs of WHOLE curves, which turns out to be an interesting metric
bound <- data.frame(do.call(rbind,sscs$hr_freqs))
library(stringr)
colnames(bound) <- str_replace(colnames(bound),"X","")
svddata <- svd(bound) 
pcdata <- prcomp(bound)

plot(y=svddata$v[,1],x=colnames(bound))
plot(y=pcdata$rotation[,1],x=colnames(bound))
plot(y=svddata$v[,2],x=colnames(bound))
plot(y=pcdata$rotation[,2],x=colnames(bound))
plot(y=svddata$v[,3],x=colnames(bound))
plot(y=svddata$v[,4],x=colnames(bound))

```

SVD is almost exactly the same
